Composition-climate paper

### 2016-02-26 Composition_Ecology.Rmd
Working on the manuscript.  Added `panderOptions`, revised the introdction.
Figure captions are all revised.  Edits throughout the methods.  Results & Discussion mostly unchanged.

Number of article references added.

---
title: "Composition-climate paper"
author: "Simon Goring *et al*."
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  word_document:
  pdf_document:
    pandoc_args: -V geometry:vmargin=1in -V geometry:hmargin=1in
  md_document:
    variant: markdown_github
csl: ecology.csl
pandoc_args: -H margins.sty
bibliography: goringetal_references.bib
---

@@

library("pander")
library("plyr")
```
-->

Abstract
====
* Ecosystems are highly dynamic and contemporary forest have a heavy imprint of land use, while climate changes over decadal and centennial scales.
* Given long time-scale climate change and continued land use conversion, are species-climate relationships today representative of those in past.
* We assess the stability of species-climate relationships using historic PLS data and FIA data.
* We show that species-climate relationships have significantly shifted for many taxa.
* Contemporary trees have shifted to cooler maximum temperatures, warmer winter temperatures, reduced seasonal temperature range, and wetter conditions.
* Relative importance of land use and climate change varies by taxon and climate variable.  
* Shifts in warmer temperature mainly due to land use change, etc.
* Implications sentence or two at the end.

@@

source('R/Setup_R.R')
source('R/load_climate.R') # gives `climate` with elements `then`, `now` and `names`
source('R/load_veg.R')

best.taxa <- c('beech', 'ironwood', 'hemlock', 'fir', 'cedar.juniper', 
              'basswood', 'spruce', 'elm', 'ash', 'pine', 'oak', 
              'tamarack', 'maple', 'poplar', 'birch')

pls_data <- get_pls('0.9-6')
agg_dens <- get_fia()[,colnames(pls_data)]

@@

source('R/natural_earth.R')

@@

Introduction
==============

Over the past 200 years centennial scale climate variability and land use change have interacted to transform the forests of the northeastern United States. There are clear indications that modern forests, as represented by the Forest and Inventory Analysis show greater homogeneity and significantly different species associations than forests surveyed as part of the Public Land Survey System, a survey that covered the western United States prior to widespread land use conversion in the late 1800s [@goring2015preset;@schulte2007homogenization].

This shift in forest composition and co-occurrence has occured over a period during which climate has begun to show the effects of anthropogenic warming, and a period in which we have increasingly used the relationship between modern tree species distributions and climate to understand future responses of tree species and forest assemblages to climate change [@iverson2013tree;@iverson1998predicting].  One of the most significant changes to forest tree distributions in the upper midwest is the loss of extensive cover through land use change [@rhemtulla2009legacies], particularly along the historical prairie and savanna margins of the region (Figure 1f), where extensive land use conversion has left little natural land.

@@

land_use$class[land_use$class == 'Water'] <- NA

@@

land_use_plot <- natural[[2]] + 
  geom_tile(data = land_use, aes(x = x, y = y, fill = class)) +
  coord_equal(xlim = ext[1:2], ylim=ext[3:4]) + 
  scale_fill_manual(values = c('#ff7f00', '#e41a1c', '#33a02c')) +
  theme(legend.position = 'none')

source('R/figures/climate_ts.R')

@@

ggsave(plot = arrangeGrob(climate_values$plot, land_use_plot, nrow=1, widths = c(3, 4)), filename='figures/figure_one.tiff')

```
![figone](figures/figure_one.tiff)

**Figure 1.** *(a-e) Climate change in the Upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions.  Low temperatures in the 1970s result in lower T~max~ during the modern era, but climate change has resulted in higher T~min~ values than the early-century normals, along with increasing P~ann~.  PDSI shows strong coherence with P~ann~ in this region. (f) Patterns of land use change in the upper Midwestern United States. Data from the NLCD [@jin2013comprehensive].*

The interaction of land use change and climatic change has been of interest to conservation managers and scientists [e.g., @pyke2004habitat] since managed natural areas often represent a biased selection from the available environmental space [@scott2001nature].  Pyke [-@pyke2004habitat] lays out a useful framework for assessing the interaction of climate and land use change.  By deliniating three types of interaction: (1) compounding, (2) confounding and (3) contradicting, it is possible to understand both the nature of the climate/land use interaction and the potential effect on apparent species resiliance and future response to broader regional change.

Compounding effects occur when both climatic change and land use change are biased in the same direction, *e.g.*, warming climate is compounded by land use change that is biased towards warmer environments.  In the case of warming, a species may lose climatic suitability along the warmest margins of its range at the same time as land use pressure increases, hastening decline in response to warming.  This pattern is seen for butterfly species in the northern Sierra Nevada mountains of California, where warming is compounded by land use conversion at the lowest elevations [@forister2010compounded].  Confounding effects may cause little overall effect on species distributions since land use pressure falls within the central distribution for the environmental gradient. Confounding effects may increase the variability of predictive models, but do little to the overall envelope.  Garry oak (*Quercus garryana*) distributions in the Pacific Northwest may suffer from confounding effects.  The spatial extent of the species has likely changed little, but it has been excluded from the deep soil sites it formerly occupied as a result of land use conversion for urban development and agriculture [@macdougall2004defining;@pellatt2014environmental], thus predictive models [@pellatt2012using] may show greater uncertainty than if complete historical distributions had been used. Contradicting effects result in a narrower climate niche for taxa, where warming forces the distribution mean higher as populations in one tail of the climate distribution experience greater environmental pressure, and, secondarily environmentally biased land use pressure in the other end of the tail excludes the taxa from those regions.  Here we might think of downward migrations of some bird species in the Sierra Nevadas in response to shifts in precipitation [@tingley2012push], where downward migration may be limited by land use conversion in the lowlands.

@@

With respect to precipitation, particularly toward the forest-prairie margin, where land use conversion continues [@wright2013recent] we would expect to see compounding effects of land use and climate change. Northern species should be shifting away from drier habitat while land use in the northern prairie-forest margin intensifies, shifting species further than expected along the precipitation gradient.

For southern tree species, land use conversion is primarily, and has primarily, occured along the southern margins of the region, parallel to temperature and precipitation gradients. This should result in compounding land use effects for all taxa with distributions in the region, resulting in narrower species distributions due to the loss of suitability and land use conversion along the southern margin of the species' range.

Critically, given the importance of modern distributional datasets for predicting species range shifts, contradicting and confounding land use effects may result in the impression of narrower climatic tolerances, or greater predictive uncertainty. Tree species are responding to climate change within forested landscapes, but the response is inconsistent. Differences in the distributions of mature trees and seedlings show responses in western North America [@monleon2015evidence], but apar to indicate range contraction, rather than northward expansion in eastern temperate North America [@zhu2012failure]. Modern correlational analysis leaves out historical distributions that can further improve our ability to predict future distributions and management implications of species shifts, potentially clouding our interpretation of these patterns. For species that reach their northern or western range limits in the upper Midwestern United States (e.g., hemlock, beech, &tc.) this may be critical, since a narrowing of the apparent environmental envelope, dut to the twin action of land use change and changing climate, may result in underprediction of species resiliance to climate change.

@@

Methods
===============
We pair pre-settlement era vegetation data from the Public Land Survey System (PLSS) aggregated to a 64 $km^{2}$ grid [@goring2015preset] with annually resolved climate data for the region.  The PLSS era data is aggregated from survey data collected on a 1 x 1 mi grid across the Upper Midwest during the PLS, from 1830 - 1904, in a time transgressive manner, from the south east to the north west.  At each survey point surveyors noted the closest two to four trees, recorded their distances to the plot center point, diameters and the azimuth to the tree.  Goring et al. [-@goring2015preset] aggregated this data to an 8 x 8 km grid to reduce local-scale variability, resulting in a gridded data product with an average of 61 plots per cell, or approximately 120 trees per cell.  When examining relationships between FIA and PLS data, Goring et al. [-@goring2015preset] report significant relationships within the FIA relating dissimilarity from PLSS era forests and the number of FIA plots within an 8x8km cell, however this relationship accounts for only 3% of the total change in dissimilarity, indicating statistical significance, but little ecological significance on this regional scale.

Another complication is the structure of the FIA dataset.  While the PLS data was collected across the western US on a uniform grid the FIA data is only collected from forested or partially forested landscapes [@woudenberg2010forest].  As such trees may be present in regions without permanant or temporary FIA plots, but no record of them exists within the FIA. This complicates our ability to assign 'absence' within the FIA, however the practice of using FIA data as a surrogate for presence/absence data is widespread [@iverson2013tree;@iverson1998predicting], and, at a broad scale, the presence of individual and isolated trees in unforested agricultural landscapes may be significant for adaptation, but it may also be the case that these trees remain on the landscape through anthropogenic agency, for example, as a result of irrigation or landscaping care within urban or rural locations, agency that ultimately leads urban ecosystems in deserts to closely resemble those in the northeastern United States [@groffman2014ecological].

@@

The choice of a thirty year normal for the PLS data is potentially problematic.  The time transgressive nature of the PLS Survey means that some of the PLS data was sampled in the early 1800s, while other come from the early 1900s [@goring2015preset;@white1983history], and for much of this time period no climatological records exist. Regardless, throughout this period climate variability exists (Figure 1d), and trees that were on the landscape from the 1830s to the 1880s are not expected to have been completely extirpated by annual scale climate variability in that same time period. As such the use of a 'pre-settlement' climate normal from 1895-1925 is likely a reasonable choice.  Regional-scale reconstructions from tree-rings and from pollen-based climate reconstructions tell us that there has been a trend of increasing temperatures (approximately 0.3^o^C in the Northern Hemisphere) since the 1800s, with pollen-based reconstructions showing even higher rates of change in the Midwestern United States [@viau2006millennial], with tree rings showing cooling at the 1800s to warming from the 1900s [@ahmed2013continental].  A second issue is that climate normals may be insufficient for modeling seedling recruitment [@serra2015averaged].  Given the widespread deforestation following settlement, one or few years of cold or hot weather could affect establishment regionally, resuling in a significantly changed landscape.

@@

source('R/figures/gain_loss_plot.R')

@@

loss_table[,1:3] <- round(tran(loss_table[,1:3], method = 'proportion') * 100,1)

ggsave(plot = loss_melt[[2]], filename = 'figures/Figure_panelloss.tiff',
       dpi=300, width=6, height = 3)
```

@@

```{r table1, results='as.is', echo=FALSE, eval=TRUE}

pandoc.table(loss_table, justify = "left")
```

```{r densities, warning=FALSE, message=FALSE, echo=FALSE, dpi=150}
source('R/resampling_climate.R')
vegclim_table <- resample_climate(climate, agg_dens, pls_data)

source('R/figures/climate_boxplots.R')
boxplot_clim <- clim_boxplots(clim_table = vegclim_table)

@@

```{r panel_densities, warning=FALSE, message=FALSE, echo=FALSE}
source('R/figures/FigureFour_Panels.R')
```

@@

source('R/hellinger_dists.R')

@@

```{r tests, warning=FALSE, message=FALSE, echo=FALSE, results='as-is', eval=TRUE}
source('R/confounding_table.R')

@@

pander::pandoc.table(conf_table, justify = "left")

@@

Our results show that the largest class of land use/climate interaction (Table 2) since the onset of EuroAmerican settlement in the Upper Midwest is Compounding, accounting for `r table(unlist(conf_table[,2:5]))['+']/(nrow(conf_table)*4) * 100`% of the possible interactions.  Confounding interactions account for `r table(unlist(conf_table[,2:5]))['.']/(nrow(conf_table)*4)*100`% of interactions, while only `r table(unlist(conf_table[,2:5]))['-']/(nrow(conf_table)*4)*100`% of interactions are contradicting.

>>>>

---
title: "Composition-climate paper"
author: "Simon Goring *et al*."
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  pdf_document:
    pandoc_args: -V geometry:vmargin=1in -V geometry:hmargin=1in
  md_document:
    variant: markdown_github
  word_document: null
csl: ecology.csl
pandoc_args: -H margins.sty
bibliography: goringetal_references.bib
---

@@

library("pander")
library("plyr")

panderOptions('missing', '--')
panderOptions('digits', 4)
panderOptions('keep.trailing.zeros', TRUE)
panderOptions('table.continues', NULL)
panderOptions('table.split.table', Inf)
pander::panderOptions('table.style', 'rmarkdown')
grDevices::ps.options(fonts = 'serif')

```
-->

Abstract
====
* Observational datasets of contemporary species-climate relationships are widely used understand environmental controls on species distriutions and project the responses of species to 21^st^-century climate change.
* Given recent climate change and continued land use conversion, are species-climate relationships today representative of those in past.
* We assess the stability of species-climate relationships using historic PLSS data and FIA data.
* We show that species-climate relationships have significantly shifted for many taxa.
* Contemporary trees have shifted to cooler maximum temperatures, warmer winter temperatures, reduced seasonal temperature range, and wetter conditions.
* Relative importance of land use and climate change varies by taxon and climate variable.  
* Shifts in warmer temperature mainly due to land use change, etc.
* Implications sentence or two at the end.

@@

source("R/Setup_R.R")
source("R/load_climate.R") # gives `climate` with elements `then`, `now` and `names`
source("R/load_veg.R")

best.taxa <- c("beech", "ironwood", "hemlock", "fir", "cedar.juniper",
              "basswood", "spruce", "elm", "ash", "pine", "oak",
              "tamarack", "maple", "poplar", "birch")

pls_data <- get_pls("0.9-6")
agg_dens <- get_fia()[,colnames(pls_data)]

@@

source("R/natural_earth.R")

@@

Introduction
==============
4)	Some researchers have focused on historical resurveys to assess climate-driven range shifts (Moritz, Waller, Tingley, etc.)
5)	Others have examined stability or plasticity of species-climate distributions during the last deglaciation (Pearman, Veloz, AT Petersen, etc.)

Global change ecologists and biogeographers use contemporary distributions of species, and, increasingly, long-term datasets to map distributions of species and species traits within environmental space.  Distributions of species within climate space can be used to assess species risk and resilience to climate change [@pellatt2012using;@iverson2013tree;@iverson1998predicting;@anderson2013nichemodels] and these models are then used to inform management and adaptation decisions in the future, although their role in developing planning scenarios is often unclear [@guisan2013predicting]. Understanding climate-related risk and response for ecological communities within landscapes is critically important for future adaptation planning, but we are often limited by a narrow window of observation that may obscure greater plasticity in species-climate responses [@maiorano2013building], and the constraints that result from historical contingency [@svenning2015influence]. Process-based species distribution models using physiological data as a primary basis [e.g., @leiblein2015modelling] may address issues of climate disequilibrium within models, focusing on representing the true niche space using experimental data at the individual and community level, however, parameterization of these models is critical, and has been shown to provide poor match at high levels [@matthes]. Increasing recognition that historical distributions show evidence of lags in species climate responses [@webb1986vegetation;@ordonez2013realized;svenning2015influence] and that habitat loss through human agency can impact apparent species distribtutions at both long [@ivory2015] and shorter time scales [@sarmento2013impacts] point to the need to understand species-climate distributions across time-scales, wherever the model lies along the gradient from correlational to process based [@dormann]and to quantify land use interactions with distribution modelling.

Focus on past distributions has taken multiple approaches, including historical resurveys [@li2014drivers;@tingley2012push] which provide evidence of sustained shifts in distribution and co-occurence [@goring2016presettlement]. Integration of paleoecological data can provide greater insight into shifting ecological patterns within communities [@jackson2015community], but one of the largest shifts in community composition within the upper Midwestern United States occured with EuroAmerican Settlement in the 1800s [@goringetal2016;@otherrefs] as a result of changing land use, and simultaneously, regional shifts in climate at the end of the Little Ice Age [@paquette2013climatic;others]. Both land use change and climate change are ongoing, and the challenge of disentangling the influence of each on modern species distributions is complex. The interaction of land use change and climatic change has been of interest to conservation managers and scientists [e.g., @pyke2004habitat] since managed natural areas often represent a biased selection from the available environmental space [@scott2001nature].  Studies into the interactions between land use and climate change have been largely focused on future impacts [@oliver2014interactions;@jetz2007projected], but widespread land use conversion has already occured and has had significant impacts on modern vegetation structure and composition at regional [@gehrig2007tree] and global scales [@].  A useful framework for assessing the interaction of climate and land use change considers three types of interaction: (1) compounding, (2) confounding and (3) counteracting [modified from @pyke2004habitat]. Using this framework it is possible to understand both the nature of the climate/land use interaction and the potential effect on apparent species resiliance and future response to regional climate change.

Compounding effects, when climatic change and land use change are biased in the same direction, are seen in butterfly species in the northern Sierra Nevada mountains of California, where warming, compounded by land use conversion at the lowest elevations drives species upslope more rapidly than might otherwise be expected [@forister2010compounded].  Confounding effects may cause little overall change on species distributions, land use pressure falls within the central distribution for the environmental gradient. The spatial extent of Garry oak (*Quercus garryana*) in the Pacific Northwest has changed little, but land use conversion has excluded it from the deep soil sites it formerly occupied [@macdougall2004defining;@pellatt2014environmental], thus predictive models for Garry oak [@pellatt2012using] may show greater uncertainty than if complete historical distributions had been used. Contradicting effects result in a narrower climate niche for taxa. Some bird species in the Nierra Nevadas should show downslope migration in response to shifts in precipitation [@tingley2012push], but this downslope migration is limited by land use conversion in the lowlands.

Over the past 200 years centennial scale climate variability and land use change have interacted to transform the forests of the northeastern United States. There are clear indications that modern forests, as represented by the Forest and Inventory Analysis show greater homogeneity and significantly different species associations than forests surveyed as part of the Public Land Survey System, a survey that covered the western United States prior to widespread land use conversion in the late 1800s [@goring2015preset;@schulte2007homogenization].  Shifts in forest composition and co-occurrence has occured over a period during which climate has begun to show the effects of anthropogenic warming, and a period in which we have increasingly used the relationship between modern tree species distributions and climate to understand future responses of tree species and forest assemblages to climate change [@iverson2013tree;@iverson1998predicting].  One of the most significant changes to forest tree distributions in the upper midwest is the loss of extensive cover through land use change [@rhemtulla2009legacies], particularly along the historical prairie and savanna margins of the region (Figure 1f), where extensive land use conversion has left little natural land.

@@

land_use$class[land_use$class == "Water"] <- NA

@@

land_use_plot <- natural[[2]] + 
  geom_tile(data = land_use, aes(x = x, y = y, fill = class)) +
  coord_equal(xlim = ext[1:2], ylim=ext[3:4]) + 
  scale_fill_manual(values = c("#ff7f00", "#e41a1c", "#33a02c")) +
  theme(legend.position = "none")

source("R/figures/climate_ts.R")

@@

ggsave(plot = arrangeGrob(climate_values$plot, land_use_plot, nrow=1, widths = c(3, 4)), filename="figures/figure_one.tiff")

```
![figone](figures/Fig_1_Composite.png)

**Figure 1.** *(a-e) Climate change in the Upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions.  Low temperatures in the 1970s result in lower T~max~ during the modern era, but climate change has resulted in higher T~min~ values than the early-century normals, along with increasing P~ann~.  PDSI shows strong coherence with P~ann~ in this region. (f) Patterns of land use change in the upper Midwestern United States. Land use data from the NLCD [@jin2013comprehensive].*

@@

With respect to precipitation, particularly toward the forest-prairie margin, where land use conversion continues [@wright2013recent] we would expect to see compounding effects of land use and climate change. Northern species should be shifting away from drier habitat while land use in the northern prairie-forest margin intensifies, shifting species further than expected along the precipitation gradient. For southern tree species, land use conversion is primarily, and has primarily, occured along the southern margins of the region, parallel to temperature and precipitation gradients. This should result in compounding land use effects for all taxa with distributions in the region, resulting in narrower species distributions due to the loss of suitability and land use conversion along the southern margin of the species' range.

Critically, given the importance of modern distributional datasets for predicting species range shifts, contradicting and confounding land use effects may result in the impression of narrower climatic tolerances, or greater predictive uncertainty. Tree species are responding to climate change within forested landscapes, but the response is inconsistent. Differences in the distributions of mature trees and seedlings show responses in western North America [@monleon2015evidence], but appear to indicate range contraction, rather than northward expansion in eastern temperate North America [@zhu2012failure]. Modern correlational analysis leaves out historical distributions that can further improve our ability to predict future distributions and management implications of species shifts, potentially clouding our interpretation of these patterns. For species that reach their northern or western range limits in the upper Midwestern United States (*e.g.*, hemlock, beech, &tc.) this may be critical, since a narrowing of the apparent environmental envelope, due to the twin action of land use change and changing climate, may result in underprediction of species resiliance to climate change.

@@

Methods
===============
Pre-settlement era vegetation data from the Public Land Survey System (PLSS), aggregated to a 64 $km^{2}$ grid [@goring2016preset] is paired with annually resolved climate data for the region.  The PLSS era data is aggregated from survey data collected on a 1 x 1 mi grid across the Upper Midwest during the PLSS, from 1830 - 1904, in a time transgressive manner, from the south east to the north west.  At each survey point surveyors noted the closest two to four trees, recorded their distances to the plot center point, diameters and the azimuth to the tree.  Goring et al. [-@goring2015preset] aggregated this data to an 8 x 8 km grid to reduce local-scale variability, resulting in a gridded data product with an average of 61 plots per cell, or approximately 120 trees per cell.  When examining relationships between FIA and PLSS data, Goring et al. [-@goring2015preset] report significant relationships within the FIA relating dissimilarity from PLSS era forests and the number of FIA plots within an 8x8km cell, however this relationship accounts for only 3% of the total change in dissimilarity, indicating statistical significance, but little ecological significance on this regional scale.

Another complication is the structure of the FIA dataset.  While the PLSS data was collected across the western US on a uniform grid the FIA data is only collected from forested or partially forested landscapes [@woudenberg2010forest].  As such trees may be present in regions without permanant or temporary FIA plots, but no record of them exists within the FIA. This complicates our ability to assign 'absence' within the FIA, however the practice of using FIA data as a surrogate for presence/absence data is widespread [@iverson2013tree;@iverson1998predicting], and, at a broad scale, the presence of individual and isolated trees in unforested agricultural landscapes may be significant for adaptation, but it may also be the case that these trees remain on the landscape through anthropogenic agency, for example, as a result of irrigation or landscaping care within urban or rural locations, agency that ultimately leads urban ecosystems in deserts to closely resemble those in the northeastern United States [@groffman2014ecological].

@@

The choice of a thirty year normal for the PLSS data is potentially problematic.  The time transgressive nature of the PLSS Survey means that some of the PLSS data was sampled in the early 1800s, while other come from the early 1900s [@goring2015preset;@white1983history], and for much of this time period no climatological records exist. Regardless, throughout this period climate variability exists (Figure 1d), and trees that were on the landscape from the 1830s to the 1880s are not expected to have been completely extirpated by annual scale climate variability in that same time period. As such the use of a 'pre-settlement' climate normal from 1895-1925 is likely a reasonable choice.  Regional-scale reconstructions from tree-rings and from pollen-based climate reconstructions tell us that there has been a trend of increasing temperatures (approximately 0.3^o^C in the Northern Hemisphere) since the 1800s, with pollen-based reconstructions showing even higher rates of change in the Midwestern United States [@viau2006millennial], with tree rings showing cooling at the 1800s to warming from the 1900s [@ahmed2013continental].  A second issue is that climate normals may be insufficient for modeling seedling recruitment [@serra2015averaged].  Given the widespread deforestation following settlement, one or few years of cold or hot weather could affect establishment regionally, resuling in a significantly changed landscape.

@@

source("R/figures/gain_loss_plot.R")

@@

loss_table[,1:3] <- round(tran(loss_table[,1:3], method = "proportion") * 100,1)

ggsave(plot = loss_melt[[2]], filename = "figures/Figure_panelloss.tiff",
       dpi=300, width=6, height = 3)
```

@@

```{r table1, results="as.is", echo=FALSE, eval=TRUE}

pandoc.table(loss_table, justify = "left", style = "rmarkdown")
```

```{r densities, warning=FALSE, message=FALSE, echo=FALSE, dpi=150}
source("R/resampling_climate.R")
vegclim_table <- resample_climate(climate, agg_dens, pls_data)

source("R/figures/climate_boxplots.R")
boxplot_clim <- clim_boxplots(clim_table = vegclim_table)

@@

```{r panel_densities, warning=FALSE, message=FALSE, echo=FALSE}
source("R/figures/FigureFour_Panels.R")
```

@@

source("R/hellinger_dists.R")

@@

```{r tests, warning=FALSE, message=FALSE, echo=FALSE, results="as-is", eval=TRUE}
source("R/confounding_table.R")

@@

pander::pandoc.table(conf_table, justify = "left", style = "rmarkdown")

@@

Our results show that the largest class of land use/climate interaction (Table 2) since the onset of EuroAmerican settlement in the Upper Midwest is Compounding, accounting for `r table(unlist(conf_table[,2:5]))["+"]/(nrow(conf_table)*4) * 100`% of the possible interactions.  Confounding interactions account for `r table(unlist(conf_table[,2:5]))["."]/(nrow(conf_table)*4)*100`% of interactions, while only `r table(unlist(conf_table[,2:5]))["-"]/(nrow(conf_table)*4)*100`% of interactions are contradicting.
